---
title: "2024-07_yang_bone_marrow_clustering"
output: github_document
---
<body style="background-color:FloralWhite;">

# BM HSC Clustering Analysis


```{r setup, message=F, warning=F}
# install.packages("Seurat")
# install.packages("remotes")
# install.packages("BiocManager")
# BiocManager::install(version = '3.18')
# BiocManager::install("glmGamPoi")
# remotes::install_github("stephenturner/annotables")
# install.packages("glmGamPoi")
# BiocManager::install("DESeq2")
# # BiocManager::install("MAST") # <- Doesn't work
# remotes::install_github("RGLab/MAST")
# # install.packages("data.filt")
# remotes::install_github("sonejilab/cellexalvrR") # <- Something Yang wanted to try
# remotes::install_github("mojaveazure/seurat-disk")

library(Seurat)
library(dplyr)
library(data.table)
library(Matrix)
library(ggplot2)
library(scales) # to better customize Seurat's plots
library(ggpubr)
library(ggrepel)
library(patchwork)
library(cowplot)
library(ggbeeswarm) # for better positioning of arrows on log2 change plots - position_jitter doesn't allow to mark points
library(future)
library(annotables) # for turning Ensembl ID to symbol
library(sctransform) # for normalization
library(glmGamPoi) # for SCTransform
library(svglite) # for vectorized, lightweight plotting
library(systemfonts) # to set the font for svg outputs
# library(DESeq2)
library(MAST)
# library(cellexalvrR) <- Something Yang wanted to try
library(SeuratDisk)

"%notin%" <- Negate("%in%")
"%notlike%" <- Negate("%like%")


# set the theme for plotting (Seurat uses ggplot's themes)
theme_set(new = theme_classic())
theme_update(
  axis.text.x = element_text(vjust = 0.5),
  strip.background = element_rect(fill = '#FFFFFF'),
  plot.title = element_text(hjust = 0.5, size = 25),
  plot.subtitle = element_text(size = 20, hjust = .5),
  axis.title = element_text(size = 23),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 18),
  legend.key.size = unit(2, 'line'),
  legend.title = element_text(size = 20, hjust = .5, vjust = .5)
  # text = element_text(family= "mono")
)

# That's not necessary (rmarkdown sets its directory as the one the .Rmd file is in.)
wd <- "/home/radgro/projects/2023-09_Yang_hsc/2024-07_bone_marrow/analysis/"
knitr::opts_knit$set(root.dir = wd)

fonts <- list(
  mono = "Consolas",
  sans = "Consolas"
)

# set svglite as a default for all the plots
# knitr::opts_chunk$set(knitr.chunk.dev = 'svglite')
# knitr::opts_chunk$set(dev = 'svglite', system_fonts = fonts)
knitr::opts_chunk$set(dev = 'svglite', dev.args = list(system_fonts = fonts),
                      cache.path = "2024-07_bone_marrow_clustering/gfm", cache = F,
                      cache.lazy = FALSE) # cache of a github_document doesn't work if the path to the gfm folder is not provided!!!

# knitr::opts_chunk$set(cache.extra = 1) # RESETS CACHE

# plan("multicore", workers = 8) # Not allowed on the server
# plan()

options(future.globals.maxSize = 8000 * 1024^2) # for seurat's parallelization

```



### Load data

```{r}
hsc_bm <- LoadH5Seurat("hsc_bm_qc.h5Seurat")
```

## Clustering analysis - continuation

The analysis, looking into different HSC markers, was started with the QC.



#### Elbow Plot
Around 10 top PCAs should be enough to obtain proper clustering.
```{r,  fig.width= 7, fig.height= 5, cache=F}
ElbowPlot(hsc_bm)
```


### Clustering 

```{r clustering_pca, fig.height=10, fig.width=15, cache = TRUE}
hsc_bm <- FindNeighbors(hsc_bm, reduction = "pca", verbose = FALSE, dims = 1:15) %>%
  FindClusters(resolution = .3, verbose = FALSE, method = "igraph")
```

```{r clustering_plots, fig.height=10, fig.width=20, cache=F}
alpha_colors_1 <- hue_pal()(4) #  number of colors equal to number of patients - required to change the alpha
alpha_colors_2 <- hue_pal()(15) #  number of colors equal to number of clusters - required to change the alpha

p1 <- DimPlot(hsc_bm, group.by = "orig.ident", pt.size = 1, reduction = "umap", cols = alpha(alpha_colors_1, .6)) +
  labs(title = "donor")

p2 <- DimPlot(hsc_bm, pt.size = 1, label = T, alpha = .5, cols = alpha_colors_2, reduction = "umap") + 
  labs(title = "cluster") + 
  theme(plot.title = element_text(hjust = .5))

p1 + p2
```

```{r clustering_plots_pca, fig.height=10, fig.width=20, cache=F}
alpha_colors_1 <- hue_pal()(4) #  number of colors equal to number of patients - required to change the alpha
alpha_colors_2 <- hue_pal()(15) #  number of colors equal to number of clusters - required to change the alpha

p1 <- DimPlot(hsc_bm, group.by = "orig.ident", pt.size = 1, reduction = "pca", cols = alpha(alpha_colors_1, .6)) +
  labs(title = "donor")

p2 <- DimPlot(hsc_bm, pt.size = 1, label = T, alpha = .5, cols = alpha_colors_2, reduction = "pca") + 
  labs(title = "cluster") + 
  theme(plot.title = element_text(hjust = .5))

p1 + p2
```

### DE analysis

```{r MAST, cache = TRUE}
markers_all <- FindAllMarkers(hsc_bm, test.use = "MAST", min.pct = 0.6, verbose = T) # minimal feature frequency of 60%
```

```{r MAST_lists}
for(i in levels(markers_all$cluster)) {print(head(markers_all[markers_all$cluster == i,], 10))}
```

Plot most differentially expressed genes
```{r most_diff_mast, fig.width=20, fig.height=7.5}
setDT(markers_all)
highest_de <- markers_all[order(avg_log2FC, decreasing = T), .SD[1:5] ,by = cluster][p_val < 5e-2][order(cluster)]
lowest_de <- markers_all[order(avg_log2FC), .SD[1:5] ,by = cluster][p_val < 5e-2][order(cluster)]

most_diff_c <- merge(highest_de, lowest_de, all = T)

markers_all_plot <- markers_all[p_val == 0, p_val := markers_all[p_val != 0, min(p_val)]]
markers_all_plot <- markers_all_plot[gene %like% "DEPRECATED", gene := gsub("DEPRECATED-", "", gene)]

ggplot(markers_all_plot, aes(x = cluster, y = avg_log2FC, color = -log10(p_val))) +
  geom_point(position = position_quasirandom()) +
  geom_text_repel(most_diff_c, mapping = aes(label = gene),
                  size = 4.5, color = "red", fontface = "bold", 
                  force = 5, min.segment.length = .5,
                  segment.size = 1) +
  geom_hline(yintercept = 0) +
  scale_color_continuous(breaks = c(1, 10, 100, 250, 320)) + 
  ggtitle(label = "Fold change and p-values of differentially expressed genes among clusters", 
          subtitle = "most differentially expressed genes are labeled") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 20),
        plot.subtitle = element_text(hjust = 0.5, size = 15))
  
```

```{r cluster_heatmap1,  fig.width= 15, fig.height= 20, cache=F}
heat_cols <- hcl.colors(3, palette = "gnbu")

highest_de <- markers_all[order(avg_log2FC, decreasing = T), .SD[1:10], 
                          by = cluster
                          ][p_val < 5e-2
                            ][order(cluster)]

heatmap <- DoHeatmap(hsc_bm[, sample(colnames(hsc_bm), 30000)], group.by = "seurat_clusters",
          features = highest_de$gene, angle = 0, hjust = .5) + 
  scale_fill_gradient2("SCTransformed\n expression", low = heat_cols[1], mid = heat_cols[2], 
                       high = heat_cols[3], guide = guide_colourbar(title.position = "top")) +
  ggtitle(label = "Expression of top 10 markers of each cluster") +
  guides(colour = "none") +
  theme(axis.text.y = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 20, vjust = .5)
        )

# heatmap[[1]]$layers[[3]]$geom_params$ymin <- heatmap[[1]]$layers[[3]]$geom_params$ymin - 2
# heatmap[[1]]$layers[[3]]$geom_params$ymax <- heatmap[[1]]$layers[[3]]$geom_params$ymax - 2
# heatmap[[1]]$layers[[4]]$aes_params$y <- heatmap[[1]]$layers[[4]]$aes_params$y -2

heatmap
```

## GOI Feature Plots

LMP: IGLL1, TOP2A,
CMPs:IL3RA, CPA3, ANXA1, 
GMPs1: S100A9, CST3, LYZ, 
GMPs2: MPO, AZU1, ELANE, PRTN3, DLK1, MS4A3, CTSG
MKPs: ITGA2B, GP9, PF4, 
MEP: TFRC, ENG, GATA1, GATA2, KLF1, GYPA, HBB, HBA1, HBA2, 
Cycling related genes: MIK67, AURKB 
HSCs: PROM1, CD34, MEIS1, MLLT3, PLEK, BLVRB, 
Others: ADGRE1, CD169 (SIGLEC1)

```{r integrins_of_interest, fig.width= 20, fig.height= 15, cache=F}
goi_list <- list("LMP" = c("IGLL1", "TOP2A"),
                 "CMPs" = c("IL3RA", "CPA3", "ANXA1"), 
                 "GMPs1" = c("S100A9", "CST3", "LYZ"),
                 "GMPs2" = c("MPO", "AZU1", "ELANE", "PRTN3", "DLK1", 
                             "MS4A3", "CTSG"),
                 "MKPs" = c("ITGA2B", "GP9", "PF4"),
                 "MEP" = c("TFRC", "ENG", "GATA1", "GATA2", "KLF1", "GYPA",
                           "HBB", "HBA1", "HBA2"),
                 "Cycling related genes" = c("MIK67", "AURKB"),
                 "HSCs" = c("PROM1", "CD34", "MEIS1", "MLLT3", "PLEK", "BLVRB"),
                 "Others" = c("ADGRE1", "CD169", "SIGLEC1"))

names <- names(goi_list)

plots <- lapply(names, function(x) {
              # plot GOIs
              p <- FeaturePlot(hsc_bm, features = goi_list[[x]], 
                          label = T, combine = F, alpha = .6)
              # format the plots
              p <- lapply(p, function(y) {
                # handles errors if the gene is not present
                plotted_goi <- y$labels$title
                y <- y +
                  theme_void() +
                  ggtitle(plotted_goi)  + # add tittle again after void
                  theme(plot.title = element_text(hjust = 0.5, size = 10),
                        legend.key.size = unit(1, 'line'))
                          })
              # join the plots
              p <- wrap_plots(p) + 
                plot_annotation(title = x,
                                theme = theme(
                                  plot.title = element_text(hjust = 0.5,
                                                            size = 12)))

})

sizes <- list(c(4, 10),
              c(4, 10),
              c(4, 10),
              c(8, 10),
              c(4, 10),
              c(8, 10),
              c(4, 4),
              c(8, 10),
              c(4, 10))

for (i in 1:length(plots)){
    ggsave(paste0("feature_plots_goi/goi_", names[i], "_feature_plot.png"), 
           plots[[i]], height = sizes[[i]][1], 
           width = sizes[[i]][2], scale = 1) # scale is set as it is easier
                                              # to format Seurats plots
                                              # that way
}


```














