---
title: "2023-06_Yang_hsc"
output: github_document
  # html_document:
  #   theme:
  #     bg: "#444444"
  #     fg: "#d6d6d6"
    
date: '2023-06-22'
---

<body style="background-color:FloralWhite;">

**Analysis of data obtained with 10X Chromium Fixed RNA Kit, Human Transcriptome.**
- 4 healthy donors
- cells
	1. CD34 enriched
	2. CD34+ CD45+ sorted
	3. frozen 
- 50k cells per donor to Chromium
___

## Prepare the workign enviornment:

-   install and load packages
-   set %notin% and %notlike%
    -   set ggplot's theme
-   set the working directory
-   set a plan for multithreading

```{r setup, message=FALSE, warning=FALSE}
# install.packages("Seurat")
# install.packages("remotes")
# BiocManager::install(version = '3.16')
# BiocManager::install("glmGamPoi")
# remotes::install_github("stephenturner/annotables")
# install.packages("glmGamPoi")
# BiocManager::install("DESeq2")
# BiocManager::install("MAST") <- Doesn't work
# remotes::install_github("RGLab/MAST")
# install.packages("data.filt")
# remotes::install_github("sonejilab/cellexalvrR") <- Something Yang wanted to try
# remotes::install_github("mojaveazure/seurat-disk")

library(Seurat)
library(dplyr)
library(data.table)
library(Matrix)
library(ggplot2)
library(scales) # to better customize Seurat's plots
library(ggpubr)
library(ggrepel)
library(patchwork)
library(ggbeeswarm) # for better positioning of arrows on log2 change plots - position_jitter doesn't allow to mark points
library(future)
library(annotables) # for turning Ensembl ID to symbol
library(sctransform) # for normalization  
library(glmGamPoi) # for SCTransform
# library(svglite) # for vectorized, lightweight plotting
library(systemfonts) # to set the font for svg outputs
# library(DESeq2)
library(MAST)
# library(cellexalvrR) <- Something Yang wanted to try
library(SeuratDisk)

"%notin%" <- Negate("%in%")
"%notlike%" <- Negate("%like%")


# set the theme for plotting (Seurat uses ggplot's themes)
theme_set(new = theme_classic())
theme_update(
  axis.text.x = element_text(vjust = 0.5),
  strip.background = element_rect(fill = '#FFFFFF'),
  plot.title = element_text(hjust = 0.5, size = 25),
  plot.subtitle = element_text(size = 20, hjust = .5),
  axis.title = element_text(size = 23),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 18),
  legend.key.size = unit(2, 'line'),
  legend.title = element_text(size = 20, hjust = .5, vjust = .5)
  # text = element_text(family= "mono")
)

# That's not necessary (rmarkdown sets its directory as the one the .Rmd file is in.)
wd <- "/disk2/user/radgro/projects/2023-09_Yang_hsc/"
knitr::opts_knit$set(root.dir = wd)

fonts <- list(
  mono = "Consolas",
  sans = "Lato"
)

# set svglite as a default for all the plots
# knitr::opts_chunk$set(knitr.chunk.dev = 'svglite')
# knitr::opts_chunk$set(dev = 'svglite', system_fonts = fonts)
knitr::opts_chunk$set(dev = 'pdf', dev.args = list(system_fonts = fonts),
                       cache.path = "2023-09_Yang_hsc/gfm/", cache = T) # cache of a github_document doesn't work if the path to the gfm folder is not provided!!!

# knitr::opts_chunk$set(cache.extra = 1) # RESETS CACHE

# plan("multicore", workers = 8) # Not allowed on the server
# plan()

```

### Load and prepare the data and metadata.

```{r Load and prepare the data}
# d like donor
d1 <- Read10X("data/A1/count/sample_raw_feature_bc_matrix/")
d1 <- CreateSeuratObject(d1, project = "a")
d2 <- Read10X("data/B1/count/sample_raw_feature_bc_matrix/")
d2 <- CreateSeuratObject(d2, project = "b")
d3 <- Read10X("data/C1/count/sample_raw_feature_bc_matrix/")
d3 <- CreateSeuratObject(d3, project = "c")
d4 <- Read10X("data/D1/count/sample_raw_feature_bc_matrix/")
d4 <- CreateSeuratObject(d4, project = "d")

```


## Merging without integration

```{r, merge_the_objects_without_integration}
hsc_pbmc <- merge(d1, y = d2)
hsc_pbmc <- merge(hsc_pbmc, y = d3)
hsc_pbmc <- merge(hsc_pbmc, y = d4)

# required in Seurat 5
hsc_pbmc <- JoinLayers(hsc_pbmc)

rm(list = c("d1", "d2", "d3", "d4"))
suppressMessages(gc())
```

## QC and filtering of cells
### Plotting
#### Main QC plots

#### Initial removal of empty droplets
```{r qc_with_empty_droplets, fig.width=10, fig.height=5, cache=F}

VlnPlot(hsc_pbmc, features = c('nCount_RNA','nFeature_RNA'), split.by = "orig.ident", pt.size = 1, raster = F, log = T) +  NoLegend()

```



```{r empty_droplets}
print("nCount_RNA")
length(WhichCells(hsc_pbmc, expression = nCount_RNA > 900))
length(WhichCells(hsc_pbmc, expression = nCount_RNA < 900))


print("nFeature_RNA")
length(WhichCells(hsc_pbmc, expression = nFeature_RNA > 900))
length(WhichCells(hsc_pbmc, expression = nFeature_RNA < 900))

```


```{r empty_droplet_removal}
hsc_pbmc <- subset(x = hsc_pbmc, subset = nCount_RNA > 900 & nFeature_RNA > 900)

```

#### Other features
```{r percentage_features}
hsc_pbmc <- PercentageFeatureSet(hsc_pbmc, pattern = "^MT-", col.name = "percent_mt")
hsc_pbmc <- PercentageFeatureSet(hsc_pbmc, "^RP[SL]", col.name = "percent_ribo")
hsc_pbmc <- PercentageFeatureSet(hsc_pbmc, "^HB[^(P)]", col.name = "percent_hb")
hsc_pbmc <- PercentageFeatureSet(hsc_pbmc, "PECAM1|PF4", col.name = "percent_plat")
```



```{r qc_main, fig.width=10, fig.height=10, cache=F}
VlnPlot(hsc_pbmc, features = c('nCount_RNA','nFeature_RNA', 'percent_mt', 'percent_hb', "percent_ribo", "percent_plat"), split.by = "orig.ident", pt.size = .5, raster = F) +  NoLegend()

```

#### Filtering
```{r feature_filtering}
hsc_pbmc <- subset(x = hsc_pbmc, subset = 100000 > nCount_RNA & percent_mt < 8)
```

``` {r umi-features, fig.width = 10, fig.height = 5, cache=F}
FeatureScatter(hsc_pbmc, "nCount_RNA", "nFeature_RNA", pt.size = 1, plot.cor = T) + scale_x_continuous(labels = scales::scientific) + NoLegend()
```


``` {r most_expressed, fig.width = 10, fig.height = 7}
counts_cells <- hsc_pbmc@assays$RNA@counts
counts_cells <- Matrix::t(Matrix::t(counts_cells)/Matrix::colSums(counts_cells)) * 100

most_expr_cells <- order(apply(counts_cells, 1, median), decreasing = T)[20:1]
most_expr_counts_cells <- as.matrix(t(counts_cells[most_expr_cells,]))

rm(list = c("counts_cells", "most_expr_cells"))
par(mar=c(5, 5, 1, 1))
boxplot(most_expr_counts_cells, cex = 1, las = 1, xlab = "% total count per cell",
        col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)

```


## Save the processed HSCs dataset
```{r save the data}
SaveH5Seurat(hsc_pbmc, filename = "data/hsc_pbmc_qc.h5Seurat", overwrite = TRUE)
```








