---
title: "2024-01_integrative_analysis_sakurai"
output: github_document
---

## Description

Seurat data integration of healthy adult donor CD34 enriched PBMCs with data from Sakurai et al. (https://www.nature.com/articles/s41586-023-05739-9).


## Setup

```{r setup, message=FALSE, warning=FALSE}
# install.packages("Seurat")
# install.packages("remotes")
# BiocManager::install(version = '3.16')
# BiocManager::install("glmGamPoi")
# remotes::install_github("stephenturner/annotables")
# install.packages("glmGamPoi")
# BiocManager::install("DESeq2")
# BiocManager::install("MAST") <- Doesn't work
# remotes::install_github("RGLab/MAST")
# install.packages("data.filt")
# remotes::install_github("sonejilab/cellexalvrR") <- Something Yang wanted to try
# remotes::install_github("mojaveazure/seurat-disk")

library(Seurat)
library(dplyr)
library(data.table)
library(Matrix)
library(ggplot2)
library(scales) # to better customize Seurat's plots
library(ggpubr)
library(ggrepel)
library(patchwork)
library(ggbeeswarm) # for better positioning of arrows on log2 change plots - position_jitter doesn't allow to mark points
# library(future) # for parallelization
library(annotables) # for turning Ensembl ID to symbol
library(sctransform) # for normalization  
library(glmGamPoi) # for SCTransform
# library(svglite) # for vectorized, lightweight plotting
library(systemfonts) # to set the font for svg outputs
# library(DESeq2)
library(MAST)
# library(cellexalvrR) <- Something Yang wanted to try
library(SeuratDisk)

"%notin%" <- Negate("%in%")
"%notlike%" <- Negate("%like%")


# set the theme for plotting (Seurat uses ggplot's themes)
theme_set(new = theme_classic())
theme_update(
  axis.text.x = element_text(vjust = 0.5),
  strip.background = element_rect(fill = '#FFFFFF'),
  plot.title = element_text(hjust = 0.5, size = 25),
  plot.subtitle = element_text(size = 20, hjust = .5),
  axis.title = element_text(size = 23),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 18),
  legend.key.size = unit(2, 'line'),
  legend.title = element_text(size = 20, hjust = .5, vjust = .5)
  # text = element_text(family= "mono")
)

# That's not necessary (rmarkdown sets its directory as the one the .Rmd file is in.)
wd <- "/disk2/user/radgro/projects/2023-09_Yang_hsc/"
knitr::opts_knit$set(root.dir = wd)
# setwd(wd)

fonts <- list(
  mono = "Lato",
  sans = "Lato"
)

# set svglite as a default for all the plots
# knitr::opts_chunk$set(knitr.chunk.dev = 'svglite')
# knitr::opts_chunk$set(dev = 'svglite', system_fonts = fonts)
knitr::opts_chunk$set(dev = 'svglite', dev.args = list(system_fonts = fonts),
                      cache.path = "/disk2/user/radgro/projects/2023-09_Yang_hsc/2024-01_integrative_analysis_sakurai_files/gfm", 
                      cache = F,
                      cache.lazy = FALSE) # cache of a github_document doesn't work if the path to the gfm folder is not provided!!!

# knitr::opts_chunk$set(cache.extra = 1) # RESETS CACHE

# plan("multicore", workers = 8) # Not allowed on the server
# plan()

```



## Load data

### Our HSCs
```{r, results='hide'}
hsc_s <- LoadH5Seurat("data/hsc_final.h5Seurat")
```

### Sakurai et al.

Load the Seurat object with finalized analysis downloaded from https://ftp.ncbi.nlm.nih.gov/geo/series/GSE192nnn/GSE192519/suppl/GSE192519%5FIntegrated%5FhHSC%2Erds%2Egz.

```{r load sakurai et al. R data, cache = F}
sakurai_s <- readRDS("2024-01_integrative_analysis_sakurai/data/GSE192519_Integrated_hHSC.rds")
```


```{r load sakurai et al. - initial data after merge, cache = F, include=F, eval=F}
sakurai <- ReadMtx("2024-01_integrative_analysis_sakurai/data/soluplus_mtx/GSM5750534_Soluplus_matrix.mtx.gz",
                   "2024-01_integrative_analysis_sakurai/data/soluplus_mtx/GSM5750534_Soluplus_barcodes.tsv.gz",
                   "2024-01_integrative_analysis_sakurai/data/soluplus_mtx/GSM5750534_Soluplus_features.tsv.gz",
                   feature.column = 1)
```

```{r , include=F, eval=F}
gt <- data.table("Geneid" = rownames(sakurai)) # gene title

gt_s <- setDT(grch38[, c("ensgene", "symbol")])
colnames(gt_s)[1] <- "Geneid"

gt_join <- gt_s[gt, on = .(Geneid), mult = 'first'][symbol %in% NA | symbol == "", symbol := Geneid]
gt_sym <- gt_join[, symbol]
```

```{r , include=F, eval=F}
rownames(sakurai) <- gt_sym

rm(list = c("gt", "gt_s", "gt_sym"))
```

```{r sakurai_seurat, include=F, eval=F}
sakurai_s <- CreateSeuratObject(sakurai, project = "sakurai")

```

```{r, include=F, eval=F}
setDT(grch38[])[symbol == "MKKS"]
table(sakurai["ENSG00000125863",]> 0)
table(sakurai["ENSG00000285508",]> 0)
table(sakurai["ENSG00000285723",]> 0)

rm(sakurai)
```


## QC

```{r qc_sakurai, fig.width=7, fig.height=5, cache=F}
VlnPlot(sakurai_s, features = c('nCount_RNA','nFeature_RNA'), pt.size = 0, raster = F, log = T) +  NoLegend()
```

``` {r umi-features, fig.width = 10, fig.height = 5}
FeatureScatter(sakurai_s, "nCount_RNA", "nFeature_RNA", pt.size = 1, plot.cor = T) + scale_x_continuous(labels = scales::scientific) + NoLegend()
```


``` {r most_expressed, fig.width = 7.5, fig.height = 5, cache = T}
counts_cells <- sakurai_s@assays$RNA@counts
counts_cells <- Matrix::t(Matrix::t(counts_cells)/Matrix::colSums(counts_cells)) * 100

most_expr_cells <- order(apply(counts_cells, 1, median), decreasing = T)[20:1]
most_expr_counts_cells <- as.matrix(t(counts_cells[most_expr_cells,]))

rm(list = c("counts_cells", "most_expr_cells"))
par(mar=c(5, 6, 1, 2))
boxplot(most_expr_counts_cells, cex = 1, las = 1, xlab = "% total count per cell",
        col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)

```
### Clusters before gene removal
```{r}
DimPlot(sakurai_s, label = T)
```


## Removal of genes not present in the 10X panel used with the initial dataset - sakurai

```{r remove_not_10X, cache=F}
panel_10x <- fread("data/gene_list_10X_panel.txt", header = F)
counts_sakurai <- GetAssayData(sakurai_s, assay = "RNA") # get the matrix from the assay data
# get the indices of genes present in the 10X panel
in_panel_10x <- which(rownames(sakurai_s@assays$RNA) %in% panel_10x[, V1])

counts_sakurai <- counts_sakurai[in_panel_10x,]

# counts_sakurai <- counts_sakurai[-hsc_counts_0,]
sakurai_s <- subset(sakurai_s, features = rownames(counts_sakurai))
```

```{r how_many_genes_per_dataset}
# HSCs dataset
print("HSC dataset")
length(which(rownames(hsc_s@assays$RNA) %in% panel_10x[, V1]))
#sakurai et al.
print("sakurai dataset")
length(which(rownames(sakurai_s@assays$RNA) %in% panel_10x[, V1]))
```

### Clustering analysis after removing genes out of the 10X panel

```{r pca_sakurai_after_gene_removal, warning = F, message = F, cache = T}
sakurai_s <- SCTransform(sakurai_s, vst.flavor = "v2", verbose = FALSE) %>%
  RunPCA(npcs = 30, verbose = FALSE) %>%
  RunUMAP(reduction = "pca", dims = 1:30, verbose = FALSE)
```


**Removing of genes resulted in mapping of negligibly worse resolution.**
```{r umap_1_sakurai_after_gene_removal}
DimPlot(sakurai_s, label = T)
```


## Integration

```{r integrate donor single-cell with sakurai et al. data, cache = T, results='asis', message = F, error = F, warning = F}
integration_list <- list(sakurai_s, hsc_s)

features <- SelectIntegrationFeatures(object.list = integration_list, nfeatures = 3000)
integration_list <- PrepSCTIntegration(object.list = integration_list, anchor.features = features)
integration_anchors <- FindIntegrationAnchors(object.list = integration_list, anchor.features = features, normalization.method = "SCT")


hsc_sakurai <- IntegrateData(anchorset = integration_anchors, new.assay.name = "integrated", normalization.method = c("LogNormalize", "SCT"))
```

```{r, eval = F, include = F}
p1 <- VlnPlot(hsc_sakurai@assays$integrated, group.by = "orig.ident", log = T)
# p2 <- VlnPlot(hsc_sakurai, features = c("nCount_SCT", "nFeature_SCT"), group.by = "orig.ident")
p1
# ggarrange(p1, p2, ncol = 1)
```

```{r, fig.width=10, fig.height=7.5, cache=FALSE}
p1 <- VlnPlot(hsc_sakurai, features = c("nCount_RNA", "nFeature_RNA"), group.by = "orig.ident", 
              log = T, slot = "scale.data", pt.size = 0)
p2 <- VlnPlot(hsc_sakurai, features = c("nCount_SCT", "nFeature_SCT"), group.by = "orig.ident", 
              slot = "scale.data", pt.size = 0)

ggarrange(p1, p2, ncol = 1)
```



### Dim Reduc

```{r, cache = T, results = 'hide', message = F, error = F, warning = F}
DefaultAssay(hsc_sakurai) <- "integrated"

# hsc_sakurai <- SCTransform(hsc_sakurai, vst.flavor = "v2", verbose = FALSE) %>%
#   RunPCA( npcs = 30, verbose = FALSE) %>%
#   RunUMAP(reduction = "pca", dims = 1:30, verbose = FALSE)
hsc_sakurai <-  ScaleData(hsc_sakurai) %>%
  RunPCA(npcs = 20, verbose = FALSE) %>%
  RunUMAP(reduction = "pca", dims = 1:20, verbose = FALSE)
  
```

```{r}
ElbowPlot(hsc_sakurai, ndims = 30)
```

#### UMAP
```{r, cache = F, fig.width= 20, fig.height= 15, cache=F, eval = F, include=T}
annot_hsc_sak <- as.data.table(hsc_sakurai@meta.data$old.ident)
annot_hsc_sak[V1 %in% NA, V1 := "Yang"]
annot_hsc_sak[V1 %notin% "Yang", V1 := "Sakurai"]
hsc_sakurai@meta.data$source <- annot_hsc_sak$V1

cluster_hsc_sak <- as.data.table(hsc_sakurai@meta.data$seurat_clusters)c
cluster_hsc_sak[V1 %in% NA, V1 := "sakurai"]
hsc_sakurai@meta.data$seurat_clusters <- cluster_hsc_sak$V1

order <- unique(hsc_sakurai@meta.data$annotation)

DimPlot(hsc_sakurai, pt.size = 2, label = T, repel = T, label.size = 6, label.box = T) 
# LabelClusters(umap_hsc_sak, id =  color = unique(ggplot_build(umap_hsc_sak)$data[[1]]$colour), size = 5, repel = T,  box.padding = 1)
```
```{r, cache = F, fig.width= 15, fig.height= 7.5, cache=F}
c25 <- c(
  "dodgerblue2",
  "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F",  # lt orange
  "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown", "darkorange"
)


# Annotate Yang's HSCs
annot_hsc_sak <- as.data.table(hsc_sakurai@active.ident)
annot_hsc_sak[V1 %notin% seq(0, 9), V1 := "Sakurai"]
hsc_sakurai@meta.data$our_clusters <- droplevels((annot_hsc_sak$V1))

# annotated Sakurai's HSCs
annot_hsc_sak <- as.data.table(hsc_sakurai@active.ident)
annot_hsc_sak[V1 %in% seq(0, 9), V1 := "Yang"]
hsc_sakurai@meta.data$Sakurai_clusters <- annot_hsc_sak$V1

order <- levels(hsc_sakurai@meta.data$our_clusters)


DimPlot(hsc_sakurai, group.by = c("our_clusters", "Sakurai_clusters"), pt.size = .4, order = order,
        label = T, repel = F, label.size = 4, label.box = T,
        cols = alpha(c25, .9))
        

```


```{r}
sessionInfo()
```



```{r}
knitr::knit_exit()
```





```{r, fig.width = 7.5, fig.height=5}

gradient <- colorRampPalette(c("#E31A1C", "dodgerblue", "green"))
cols_plot <- c(gradient(10), "khaki2")

DimPlot(hsc_sakurai, group.by = c("our clusters", "Sakurai's clusters"), pt.size = 1, label = T, repel = T, label.size = 6, label.box = T, cols = alpha(c25, .7)) 
```

#### Save the processed dataset

```{r}
SaveH5Seurat(hsc_sakurai, filename = "data/hsc_sakurai.h5Seurat", overwrite = TRUE)
```
